\section{Introduction}

Advanced Persistent Threats (APTs) have become a major threat to national, enterprise, and critical-infrastructure cybersecurity. According to M-Trends 2026, the global median dwell time rose to 14 days in 2025, while the median dwell time for cyber-espionage intrusions reached 122 days\cite{mandiant2026mtrends}. Verizon DBIR 2025 reports that credential abuse and vulnerability exploitation remain major initial attack vectors, and that the share of breaches involving third parties doubled to 30\%\cite{verizon2025dbir}. IBM's 2024 Cost of a Data Breach report further shows that the global average cost of a data breach reached \$4.88 million, and that about 70\% of affected organizations experienced significant or moderate business disruption\cite{ibm2024costdatabreach}. In response, organizations deploy layered defense stacks, including Endpoint Detection and Response (EDR), Next-Generation Firewalls (NGFW), Security Information and Event Management (SIEM), Privileged Access Management (PAM), and email security gateways.

However, deployment does not imply effectiveness. A defense stack with more products is not necessarily more capable, and a system that fails to stop an entire APT campaign is not necessarily useless. APT campaigns are multi-stage, long-lived, and heterogeneous: an attacker may enter through phishing, public-facing vulnerabilities, or credential abuse; then establish command and control, collect credentials, move laterally, evade detection, and eventually exfiltrate data or cause impact. These characteristics impose three requirements on defense evaluation. First, the evaluation should distinguish where in the attack chain a defense succeeds, because blocking initial access is not equivalent to detecting exfiltration after compromise. Second, it should account for alternative and dependent attack paths: in some stages the attacker only needs one successful path, while in others all required actions must succeed. Third, it should explain which defensive control failed first, because the same missed detection may be caused by weak identity controls, insufficient isolation, missing endpoint hardening, or poor observability. Therefore, APT defense evaluation should not only answer whether an attack was stopped, but also explain which stages, control domains, and product categories are responsible for defensive success or failure.

Existing evaluation and scoring approaches only partially address this problem. MITRE ATT\&CK Evaluations provide detailed step-level observations of product behavior in adversary-emulation scenarios\cite{mitreAttackEvaluations}. Such results are valuable for understanding whether a product observed, detected, or protected against specific attack steps, but they do not directly translate those observations into stage-aware scores or explain which victim-side control point was the first failed control. Other work evaluates system security using vulnerability severity, attack difficulty, risk values, or attack-graph-based reachability\cite{mell2007common,wang2007measuring,erola2022system}. These methods support quantitative risk analysis, but their main concern is usually which vulnerability or path is risky, rather than how a deployed defense stack performs against action-level APT behaviors. As a result, three gaps remain: existing scores are often black-box summaries, attack stages are frequently treated without differentiated defensive value, and missed or weakly handled actions are rarely attributed to concrete defensive control failures.

These gaps motivate a control-point-oriented intermediate representation. Instead of scoring a defense result directly from raw attack steps, this paper first maps each attack action to the victim-side control point that failed first and is most worth repairing. This representation is inspired by the observation that the same ATT\&CK technique may correspond to different repair actions in different contexts: credential use may originate from brute force, credential dumping, hard-coded secrets, or phishing, each requiring a different defensive response. To make this idea usable for scoring, three challenges must be addressed. The attribution schema must be stable enough to classify heterogeneous APT actions; the attack-stage dimension must be modeled without forcing the control domains into a fixed kill-chain order; and action-level observations must be aggregated according to the logical structure of each stage rather than by a simple average.

To address these challenges, this paper proposes a stage-aware APT defense capability scoring framework based on control-point attribution. The framework contains three layers. First, a control-point attribution layer classifies each attack action by the victim-side first-failed control point, using 9 top-level domains and 94 leaf-level control points. Second, a benchmark construction layer enriches each action with a canonical ATT\&CK phase, a defense opportunity map, and a defense-technology crosswalk that maps control points to defense product categories. Third, a scoring engine aligns tested defense results with benchmark actions, assigns action-level hit values for \texttt{PREVENTED}, \texttt{BLOCKED}, \texttt{DETECTED}, and \texttt{MISSED} outcomes, and aggregates them into stage-level, playbook-level, phase-wise, control-domain, and product-category scores. The stage aggregation is explicitly modeled: OR stages use the weakest covered alternative, AND stages use the strongest interruption point, and Cluster stages use weighted averaging.

The main contributions of this paper are as follows:
\begin{itemize}
\item We propose a control-point attribution schema for APT defense evaluation, shifting the evaluation unit from `what the attacker did'' to `which victim-side control failed first.'' The schema covers 9 top-level domains and 94 leaf-level control points, enabling explainable attribution of defensive weaknesses.
\item We design a benchmark extension architecture that connects control-point attribution with stage-aware scoring. It includes canonical phase mapping, a defense opportunity map, and a defense-technology crosswalk, allowing defense results to be analyzed by attack phase, control domain, and product category.
\item We develop a stage-aware scoring model that separates action-level hit values from stage importance and formalizes OR, AND, and Cluster stage semantics. This avoids treating all APT steps as equally valuable and supports interpretable aggregation from actions to playbooks.
\item We evaluate the framework on 53 APT playbooks containing 1,758 attack actions. The results validate the mathematical behavior of the scoring engine and show that no single product category can independently cover the full APT attack surface, highlighting the need for coordinated defense-in-depth.
\end{itemize}

